% -*- fill-column: 80; TeX-command-extra-options: "-shell-escape" -*-
\documentclass{report}
\usepackage{minted}
\usepackage{syntax}
\usepackage{mathpartir}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikz}
\usetikzlibrary{graphs,graphdrawing,decorations.pathreplacing,decorations.pathmorphing,arrows.meta}
\usegdlibrary{layered,trees}
\usepackage{multicol}

\usepackage{fontspec}
\setmonofont{Menlo}[Scale=0.8]

\usepackage{newunicodechar}
\newfontface\mathsymbolfont{STIXGeneral}
\newunicodechar{⟦}{{\mathsymbolfont{\llbracket}}}
\newunicodechar{⟧}{{\mathsymbolfont{\rrbracket}}}
\newunicodechar{⊢}{{\mathsymbolfont{\vdash}}}
\newunicodechar{⦂}{{\mathsymbolfont{⦂}}}
\newunicodechar{⋎}{{\mathsymbolfont{\curlyvee}}}
\newunicodechar{∷}{{\mathsymbolfont{∷}}}

\begin{document}

\newcommand{\llbracket}{[\![}
\newcommand{\rrbracket}{]\!]}
\newcommand{\IO}{\mathsf{IO}}
\newcommand{\bind}{>\!\!>\!\!=} \newcommand{\concbind}{>\!\!>\!>\!\!=}
\newcommand{\subtyp}{\geq:}
\newcommand{\notsubtyp}{\ngeq:}
\newcommand{\lift}[1]{\ensuremath{\llbracket#1\rrbracket}}
\newcommand{\use}[2]{\ensuremath{\llbracket#2\rrbracket_{#1}}}
\newcommand{\ftv}{\operatorname{FTV}}
\newcommand{\fv}{\operatorname{FV}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\author{Luke Lau}
\title{A Resourceful Monad for IO}
\begin{titlepage}
  \begin{center}
    {\Huge A Resourceful  Monad for IO}
    \vskip 0.25in
    \centerline{\Large Luke Lau}
    \vfill
    \large
    {\hsize5in
      {\uppercase{\Large \bf A Dissertation}}
      \vskip 0.2in
      Presented to the University of Dublin, Trinity College \\
      In Partial Fulfilment of the Requirements for the Degree of \\
      \vskip 0.2in
      {\bf Master of Science in Computer Science}
    }

    \vfill
  \end{center}

  \noindent {\bf Supervisor of Dissertation} \\
  Glenn Strong

\end{titlepage}

\tableofcontents

\chapter{Introduction}
The immutability and purity of some functional languages make them seem like a
perfect fit for parallelism and concurrency. The lack of side effects mean we
are free to compute expressions in whatever order we please, and there is no
shared mutable state to plague us with race conditions. However in the
real world, code is never completely pure. Useful programs need to interact with
the real world at some point, whether that be by reading from a keyboard or
sending packets over a network. That is, they need to carry out side
effects: And the order in which they are carried out is important.

So when it comes to side effects in a language with referential transparency,
the language must model the sequence in which effects are carried out. One of
the most successful approaches has been to capture the state of the outside
universe, and have functions with side effects write to that state. This way, if
writing to \texttt{stdout} causes some particles to be perturbed on a display
monitor, then we can track that by saying the state of the universe has changed,
and that it matters what order it changed in. We don't actually track what
particles are there, where they might be or what their current charge is. All we
need to know is that the universe is slightly different than what it was before.

Concurrent Clean models this by threading the world in and out of
functions. Uniqueness types guarantee that the same world is only used once, so
that the programmer doesn't create an alternative timeline by duplicating it.
\begin{minted}[breaklines]{clean}
readFile :: !String !*World -> (!MaybeError FileError String, !*World)
\end{minted}
Haskell also treats the world as a state, but without the uniquness guarantee.
Any function that interacts with the world returns a function, which returns a
new version of the World alongside the function result.\footnote{The actual
  definition in GHC is \\ \mintinline{haskell}{newtype IO a = IO (State#
    RealWorld -> (# State# RealWorld, a #))}}

\mint{haskell}|type IO a = World -> (World, a)|

How Haskell ensures that an old world isn't erroneously used instead of a
fresh new one is by hiding the updating of the state from the user, tucking it
away into a \textit{monad}. This idea of using monads to sequence together
stateful computations was first introduced by Peyton Jones and
Wadler~\cite{peytonjones1993}\cite{wadler1995}. The programmer no longer needs
to keep track of the state of the universe. They can keep their imperative code
imperative, and their pure code pure. This marriage of monads and IO is one of
the crown jewels of functional language research to come out of Haskell.
\begin{minted}{haskell}
instance Monad IO where
  return x = \w -> (w, x)
  x >>= f = \w -> let (w', y) = x w in f y w'
\end{minted}
Now IO actions can be easily chained together in a type-safe way that ensures
their ordering.  Unfortunately, this ordering imposes limitations. One of the
main benefits of pure functional languages is that since expressions do not
have side effects, there is no restriction on what order they need to be
evaluated in.  Take for example the following program:
\begin{minted}{haskell}
f, g, h :: Int -> Int
f x = g x + h x
\end{minted}
\texttt{g} could be evaluated before \texttt{h}, or \texttt{h} could be
evaluated before \texttt{g}. It wouldn't make a difference because there are no
side effects. One might be tempted then to evaluate the two expressions
concurrently, and indeed that would be safe to do so.  The same cannot be said
for impure IO actions however, and Haskell's type system is well aware of that.
\begin{minted}{haskell}
f, g, h :: Int -> IO Int
f x = g x >>= \y -> y + h x
\end{minted}
We need to explicitly bind the actions and sequence evaluation.  Does this mean
that concurrency is impossible for IO actions? Not at all, many languages
provide primitives to run these actions concurrently in a type safe
way. Haskell's \texttt{base} library has \mintinline{haskell}{forkIO}, but for
simpliclty we are going to assume the existence of a higher level function that
runs two IO actions simultaneously and collects the results.
\begin{minted}{haskell}
concurrently :: IO a -> IO b -> IO (a, b)
\end{minted}
Now we can use it to run our two IO actions side by side safely. 
\begin{minted}{haskell}
f, g, h :: Int -> IO Int
f x = g x `concurrently` h x >>= \(a, b) -> return (a + b)
\end{minted}
But what if \texttt{g} and \texttt{h} actually looked like this?
\begin{minted}[samepage]{haskell}
g x = do
  txt <- readFile "foo.txt"
  return (x + (read txt))
h x = do
  writeFile "foo.txt" "hello"
  return (42 - x)

readFile :: FilePath -> IO String
writeFile :: FilePath -> String -> IO ()
\end{minted}

Running these two functions concurrently could be disastrous, as the order in
which they execute could affect the outcome of the program. These innocuous
looking IO actions then end up introducing non-determinism and race conditions.
\begin{samepage}
We know statically that a program such as
\mint{haskell}|writeFile "foo.txt" "a" `concurrently` writeFile "foo.txt" "b"|
\noindent should probably not be allowed, because it is blatantly accessing the
same resource simultaneously.
\end{samepage}
But then why did the type system allow it?  Has it failed us? The goal of the
type system is to disallow as many incorrect programs as possible whilst
allowing all correct programs. It is a fine line as to what programs are
deemed correct and what are deemed incorrect. A type system too lenient and
buggy programs will creep through: but a type system too strict and the
programmer will waste time fighting the type checker.

In this dissertation we aim to find a point in the design space that rejects
such programs as the one above. We do \textbf{not} want to allow programs that
have glaring race conditions, where we can see that there is a contentious
access of resources. Our work is based around the idea of keeping track of what
resources are in use, at the type level. We first imagine what this might look
like in Haskell by adding another type parameter to
our IO type, representing what resource an IO action uses:
\mint{haskell}|type IO r a = World -> (World, a)|
$r$ is a phantom type parameter, which only exists at the type level. Now our
type signatures could look like this, annotating the API with what resources it
might use.
\begin{minted}{haskell}
data Resource = FileSystem | Net | Database | OpenGL | ...
readFile :: FilePath -> IO FileSystem ()
writeFile :: FilePath -> IO FileSystem String
readSocket :: Socket -> IO Net ()
runQuery :: Query a -> IO Database a
swapBuffers :: IO OpenGL ()
\end{minted}

Keep in mind we are painting in broad strokes when we use the word
``resource''. In the running example the resource has been a file,
\texttt{foo.txt}, but the notion of a resource can be as broad or as specific as
a function needs it to be. It could represent a specific database
instance, or a network socket. For simplicity in our example we will
consider the entire file system as a single resource, the entire network as
a single resource, and so on.

Now that we know what resources each IO action is using, we would like to change
the type of our concurrent function to take advantage of this new
information. Perhaps we would like to reject any two functions that use the same
resource, i.e.\ it only accepts IO actions with distinct resources.
\mint{haskell}|concurrently :: r \~ s => IO r a -> IO s b -> IO ? (a, b)|
You can read \verb$r \~ s$ as ``r is distinct from s'', or the opposite of a
\verb$r ~ s$ equality constraint that one might see in a type signature. This,
however, does not exist in Haskell.  And what does it exactly mean for two
resources to ``be distinct''? And what resources would the returned $\IO$ use?

These questions are answered with a
formal definition of a type system that tracks resource usage. We explore a
specific point in the design space, where the type system rejects programs like
\[ \textsf{readFile} \curlyvee \textsf{readFile} \]
but accepts and assigns types to programs such as
\[ \textsf{readFile} \curlyvee \textsf{readNet} : \IO_{\textsf{File} \cup \textsf{Net}} \
  \square \times \square \]
In short, we aim to create a type system that keeps tracks of the resources
being used, so that the programmer doesn't have to. It \textbf{does not} aim to
solve concurrency --- there will still be programs that have concurrency errors
that the type system will still allow. We just aim to narrow down the scope of
valid programs, by ruling out those with blatant, concurrent resource access
errors. 

\section{Overview}
This dissertation has been written in a style that doesn't assume prior
knowledge of type theory, and so it should double as a tutorial along our
journey to create our resourceful type system.

In Chapter~\ref{chapter:background} we will talk about the inspirations of the
type system, namely separation logic, and how it parallels with our monadic
language. We will also briefly go over what we mean by a monad formally, and
then look at the original design of Hindley-Damas-Milner which we will build
upon.
Chapter~\ref{chapter:system} introduces the language, its type system and its
semantics.  It gives a complete definition of all the parts needed to prove
properties.  We prove these properties in Chapter~\ref{cha:properties}, in which
we eventually build up and present a proof of its soundness. This proof, and the
others that accompany it, are then mechanised within the dependently typed
language and proof assistant Agda. The methodology of this is discussed in
Chapter~\ref{cha:mechanisation}. Finally, Chapter~\ref{cha:evaluation}
evaluates our type system. We look at how it might be refined, and find many
areas of further work that would be interesting to explore.

\include{background}

\include{system}

\include{properties}

\include{mechanisation}

\include{evaluation}

\chapter{Conclusion}
We have presented a pragmatic approach to reasoning about concurrency within a
type system. We build upon the Hindley-Damas-Milner type system, incorporating
the idea of using monads for modelling effects, and extend it with our
resourceful constructs. We can prove it retains type soundness, and in the
process of doing so we have also created a framework for mechanising the proof
of HDM's soundness within Agda.

The system is simple, but it occupies a point in the design space that fits well
into existing functional programming languages, and would provide immediate
utility in keeping track of concurrent resourcefulness. Furthermore, there are
many different interesting directions the system can be extended in.

\bibliographystyle{acm}
\bibliography{report}

\include{appendix}

\end{document}

% LocalWords: stdout LocalWords Idris haskell Agda
%%% Local Variables:
%%% TeX-engine: luatex
%%% End: